GMSE: an R package for generalised management strategy evaluation

### 2017-07-17 notebook/ms/ms.Rmd
New edits to the vignette and manuscript draft

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie, Jeremy J. Cusack, Roc&#0237;o Pozo, O. Sarobidy Rakotonarivo,
  and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
linestretch: 1
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
bibliography: gmse.bib
---

@@

         ylab = "Stake-holder % yield", cex.lab = 1.25);

@@

1. Management strategy evaluation (MSE) is a powerful tool for simulating all aspects of natural resource mangement under conditions of uncertainty.
2. We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive management scenarios under complex social-ecological interactions and uncertainty.
3. GMSE is agent-based and spatially explicit, and incorporates a high degree of realism and uncertainty through mechanistic modelling of the social-ecological system.
4. We show how GMSE can be used to simulate a social-ecological system, providing the example of management of a waterfowl population that exploits agricultural land, whose conservation is there in conflict with stakeholders.
5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.

@@

Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017]. Conservation of biodiversity can be complicated by the immediate need to allocate natural resources and land area for human food production, causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a]. Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century. Effective management tools are therefore needed for the long-term maintenance of natural resources under the rising demand for food production.

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions. This approach can be further improved by considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of managers and stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions. Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of managers and stakeholders.

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stake-holder decision-making. Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique individual goals, allowing GMSE to model scenarios of conservation conflict.

@@

Game theory is the standard tool for understanding conflict bewteen rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016]. A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable. Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, replacement), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].

In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for managers or stakeholders. Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each agent to heuristically find strategies that are well-adapted to achieving agent goals. Hence for the manager and each stakeholder, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised. In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the agent's strategy in the current time step.

@@

Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of gmse() is an exhaustive list that includes all resources and observations, all manager and stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function). An example below shows how simulations are set and interprted.

@@

|  land_dim_1     |  100    | Height of the landscape (vertical cells)                                                         |

@@

|  remove_pr      |  0      | Density-indepdent probability of resource mortality during a time step                           |

@@

|  res_move_type  |  1      | Type of resource movement (default up to res_movement cells in any direction)                    |

@@

|  tend_crop_yld  |  FALSE  | Proportional increase per landscape cell from tend_crops action                                  |

@@

|  manage_freq    |  1      | Frequency (in time steps) with which managers enact policy                                       |

@@

An example of crop management
===================================================

Here we consider the example of a managed natural resources whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2017a; @Fox2016a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops. Using GMSE, we can simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring waterfowl (i.e., 'policy') for ten farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of 10 stakeholders. The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation. The upper right panel shows the same landscape broken down into 10 differently coloured regions, which correspond to areas of land owned by each of the 10 stakeholders. The middle left panel shows the actual abundance of resources (black solid line, and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals of a mark-recapture analysis), over time. The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of stakeholder performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all stakeholders over time (some actions might be unsuccessful if resources are unavailable on a stakeholder's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```

@@

The GMSE package is under further development to include additional features that will be of interest to conservation biologists, managers, and the general public. Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved manager and stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface. Additionally, future versions of GMSE will allow software users to take the place of managers or stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function). Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework. The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.

>>>>

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie, Jeremy J. Cusack, Roc&#0237;o Pozo, O. Sarobidy Rakotonarivo,
  and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
linestretch: 1
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
bibliography: gmse.bib
---

@@

         ylab = "stakeholder % yield", cex.lab = 1.25);

@@

1. Management strategy evaluation (MSE) is a powerful tool for simulating all aspects of natural resource mangement under conditions of uncertainty.
2. We present the R package GMSE, which generalises MSE using a game-theoretic approach to simulate adaptive management scenarios under complex social-ecological interactions and uncertainty.
3. GMSE is agent-based and spatially explicit, and incorporates a high degree of realism and uncertainty through mechanistic modelling of the social-ecological system.
4. We show how GMSE simulates a social-ecological system using the example of a waterfowl population that is adaptively managed; simulated waterfowl exploit agricultural land, causing conflict between conservation interests and the interest of stakeholders (farmers) in maximising crop yield.
5. The R package GMSE is open source under GNU Public License; source code and documents are freely available on GitHub.

@@

Many global natural resources, including the biodiversity on which critical ecosystem services depend, are in a state of severe decline [@Dirzo2014; @Hautier2015; @Ceballos2017; @OConnell2017]. Conservation of biodiversity can be complicated by the immediate need to use natural resources and land area for human food production, causing real or perceived conflicts between conservation and food security and creating a challenge for the management of many natural resources [@Redpath2015a]. Given an increasing human population [@Crist2017], the number and intensity of such conflicts are likely to increase into the twenty first century. Effective management tools are therefore needed for the long-term maintenance of natural resources under the rising demand for food production.

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions. This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of managers and stakeholders [@Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions. Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of managers and stakeholders.

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stakeholder decision-making. Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.

@@

Game theory is the standard tool for understanding conflict between rational agents, and is therefore the appropriate tool for modelling manager and stakeholder actions and addressing issues of cooperation and conflict in conservation [@Colyvan2011; @Lee2012; @Kark2015; @Adami2016; @Tilman2016]. A game-theoretic approach is compatible with agent-based modelling [@An2012; @Tesfatsion2017], but given the number of possible strategies available to agents, identifying optimal strategies is often intractable. Instead, genetic algorithms, which mimic the process of natural selection (mutation, recombination, selection, reproduction), are useful to find adaptive solutions for game strategies [e.g., @Balmann2000; @Tu2000; @Hamblin2013].

In keeping with the MSE approach [@Bunnefeld2011], GMSE does not attempt to find optimal strategies or solutions for managers or stakeholders. Instead, within each time step of a simulation (Figure 1), a new genetic algorithm is called for each agent to heuristically find strategies that are well-adapted to achieving agent goals. Hence for the manager and each stakeholder, a unique population with a random set of possible values affecting policy or actions, respectively, is initialised. In each generation of the genetic algorithm, strategies crossover and mutate; when this results in strategies that are over-budget, values are iteratively decreased at random until budget constraints are satisfied. A fitness function then evaluates each strategy in the population, and a tournament is used to select the next generation of strategies [@Hamblin2013]. The genetic algorithm terminates when a minimum number of generations has passed and the increase in the fitness of the fittest strategy between the current and previous generation is below some threshold. The highest fitness strategy in the population then becomes the agent's strategy in the current time step.

@@

Simulations are run using the gmse() function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of gmse() is an exhaustive list that includes all resources and observations, all manager and stakeholder decisions and actions, and all landscape properties in each time step of the simulation. Results are most easily interpreted visually, so a summary of simulation dynamics is plotted by default (the plot can also be called using the plot_gmse_results function). An example below shows how simulations are set and interpreted.

@@

|  land_dim_2     |  100    | Height of the landscape (vertical cells)                                                         |

@@

|  remove_pr      |  0      | Density-independent probability of resource mortality during a time step                         |

@@

|  res_move_type  |  1      | Type of resource movement (default is up to res_movement cells in any direction)                 |

@@

|  tend_crop_yld  |  0.2    | Proportional increase per landscape cell from tend_crops action                                  |

@@

|  manage_freq    |  1      | Frequency (in time steps) with which managers revise and enact policy                            |

@@

An example of resource management
===================================================

Here we consider the example of a managed natural resource whose abundance affects a group of stakeholders by temporarily decreasing the value of stakeholder land. This scenario could be interpreted in multiple ways; we consider a protected population of waterfowl that exploits agricultural land causing a conservation conflict with farmers [e.g., @Tuvendal2015; @Fox2016a; @Fox2017a]. Managers in this example might want to keep the abundance of waterfowl at a target level, while farmers might want to minimise the damage inflicted on their crops. Using GMSE, it is possible to simulate waterfowl population dynamics, along with the continued monitoring and policy set by managers, and the actions that farmers take to protect their crop yields given the constraints of policy. Here we consider a population of waterfowl with an initial abundance and manager target abundance of 1000, but whose carrying capacity (applied to mortality) is 2000. Waterfowl consume and destroy all crop yield upon arrival to a landscape cell. Managers estimate population size in each time step using mark-recapture techniques, and use their estimates to set costs of culling and scaring (non-lethal) waterfowl (i.e., 'policy') for ten farmers. Farmers attempt to reduce the negative impact of waterfowl on the cropland that they own, working within the constraints of culling and scaring costs and their budget for performing these actions.

@@

```{r, echo = FALSE, fig.height = 6, fig.align = 'left', fig.cap = "Results of an example simulation illustrating the management of a protected resource that exploits the land of 10 stakeholders. The upper left panel shows locations of resources (black dots) on the landscape in the final time step of the simulation. The upper right panel shows the same landscape broken down into 10 differently coloured regions, which correspond to areas of land owned by each of the 10 stakeholders. The middle left panel shows the actual abundance of resources (black solid line, and the abundance of resources as estimated by the manager (blue solid line; shading indicates 95 percent confidence intervals of a mark-recapture analysis), over time. The horizontal dotted red and blue lines show the resource carrying capacity enacted on adult mortality and the manager's target for resource abundance, respectively. The orange line shows the total percent yield of landscape cells. The middle right panel shows total percent yield of landscape cells for each individual farmer, differentiated by colour, where line colours correspond to areas of the landscape in the upper right panel. The lower left panel shows the cost of stakeholders performing actions over time, as set by the manager. The lower right panel shows the total number of actions attempted to be performed by all stakeholders over time (some actions might be unsuccessful if resources are unavailable on a stakeholder's land to cull or scare, so, e.g., culling actions might be larger than resources actually culled)."}
plot_gmse_results_vign(res = sim$resource, obs = sim$observation, 
                       land = sim$land, agents = sim$agents, paras = sim$paras,
                       ACTION = sim$action, COST = sim$cost);
```

@@

The GMSE package is under continued development to include additional features that will be of interest to conservation biologists, managers, and the general public. Such features will include multiple (interacting) resource types and sub-types (e.g., structured populations), improved manager and stakeholder decision-making and prediction based on multiple time steps of simulation history, incorporation of empirical data (e.g., landscape features), and a browser-based graphical user interface. Additionally, future versions of GMSE will allow software users to take the place of managers or stakeholders in decision-making during simulations (replacing the genetic algorithm as desired), facilitating the collection of data to test sociological hypotheses and further improve the realism of the genetic algorithm (for a preliminary example of this, see the 'hunt' argument in the gmse() function). Code underlying GMSE is publicly available on GitHub < https://github.com/bradduthie/gmse > and highly modular, meaning that GMSE model components (black boxes in Figure 1) can be developed freely and independently, then integrated into the broader GMSE framework. The GMSE R package is therefore a versatile and collaborative tool with widespread applications for adaptive resource management.
